\pkg{DNest4}: Diffusive Nested Sampling in \proglang{C++} and \proglang{Python}

### 2016-05-26 paper/ms.tex
proposal

\section{Running DNest4}\label{sec:running}
To demonstrate \pkg{DNest4},
we will use a simple linear regression example where the
sampling distribution is
\begin{align}
y_i | m, b, \sigma &\sim \textnormal{Normal}(mx_i + b, \sigma^2)
\end{align}
and the priors are
\begin{align}
m &\sim \textnormal{Normal}(0, 1000^2)\\
b &\sim \textnormal{Normal}(0, 1000^2)\\
\ln\sigma &\sim \textnormal{Uniform}(-10, 10).
\end{align}
These are naive diffuse priors and do not have any special status.
This model was implemented in \proglang{C++}
(see Section~\ref{sec:models} for details).
The dataset is shown in Figure~\ref{fig:regression_lines}
and the code is included in the
{\tt code/Examples/StraightLine} subdirectory. A simplified version of the
code is explained in Section~\ref{sec:models}.

To execute \pkg{DNest4} on this problem, run the executable {\tt main}
in the {\tt code/Examples/StraightLine} directory. The output to the screen
should contain information about levels and ``saving particles to disk''.
After 5000 particles have been saved to the disk, the run should terminate.

@@

When implementing a model class, it is imperative that the
\code{from\_prior(DNest4::RNG\&)} and \code{perturb(DNest4::RNG\&)}
member functions be consistent with each other. One technique
for verifying this is to sample the prior for a long time
(by setting the maximum number of levels to 1) and inspect
{\tt sample.txt} to ensure that each parameter is exploring
the correct distribution.

>>>>

\section{Running DNest4}\label{sec:running}
To demonstrate \pkg{DNest4},
we will use a simple linear regression example where the
sampling distribution is
\begin{align}
y_i | m, b, \sigma &\sim \textnormal{Normal}(mx_i + b, \sigma^2)
\end{align}
and the priors are
\begin{align}
m &\sim \textnormal{Normal}(0, 1000^2)\\
b &\sim \textnormal{Normal}(0, 1000^2)\\
\ln\sigma &\sim \textnormal{Uniform}(-10, 10).
\end{align}
These are naive diffuse priors and do not have any special status.
This model was implemented in \proglang{C++}
(see Section~\ref{sec:models} for details).
The dataset is shown in Figure~\ref{fig:regression_lines}
and the code is included in the {\tt code/Examples/StraightLine}
subdirectory. A simplified version of the
code is explained in Section~\ref{sec:models}.
To execute \pkg{DNest4} on this problem, go to this directory and
execute {\tt main}. The output to the screen
should contain information about levels and ``saving particles to disk''.
After 5000 particles have been saved to the disk, the run should terminate.

@@

Recall that for Nested Sampling, the Metropolis acceptance probability,
excluding the term for the likelihood, is
\begin{align}
H = \frac{q(\params'|\params)}{q(\params | \params')}
\times \frac{\pi(\params')}{\pi(\params)}.
\end{align}
When implementing a model class, the
\code{perturb(DNest4::RNG\&)}
member function must return the logarithm of this value.
Since the prior for $m$ is a normal distribution with
mean zero and standard deviation $1000$, $H$ is
\begin{align}
H &= \frac{\exp\left[-\frac{1}{2}(m'/1000)^2\right]}
{\exp\left[-\frac{1}{2}(m/1000)^2\right]}.
\end{align}
This explains the use of the \code{log\_H} variable.

It is imperative that \code{from_prior} and
\code{perturb}
be consistent with each other. One technique
for verifying this is to sample the prior for a long time
(by setting the maximum number of levels to 1) and inspect
{\tt sample.txt} to ensure that each parameter is exploring
the prior correctly.
